DRAFT \\ From physical principles to classical and quantum \\ Hamiltonian and Lagrangian particle mechanics

### 2016-11-18 2015CompleteDerivation/completeDerivation.tex
Adding Christine's comments

We will see that, as a first approximation, quantum particles are more keenly described by classical composite states (i.e. distributions in phase space) like any other object made of a finite amount of material. We'll also see that classical distributions already possess qualitative properties that are typically associated with quantum systems, including a classical analogue of the uncertainty principle.

@@

The first one is that quantum particles are not associated with a well defined value for position and momentum. If we misidentify the electron with a classical particle, this would indeed be worth of note and puzzlement. But if the electron is a classical distribution, there is no issue. The earth is both under Milano and New York. Your right index finger is not, hopefully, at the same position of your left eye. In fact, based on what we saw before, any object of finite mass is a distribution over a region of phase space, and discontinuous distributions are ruled out by \ref{continuous_map}. An electron is an object of finite mass: why should it be any different? In fact, if one claimed an electron to be point-like, he would be claiming an electron behaves in a different way than all other physical objects of finite mass/energy (which are the only ones we experimentally work with). Shouldn't \emph{he} have the burden of proof?

@@

While this picture is quite reasonable, some people may reject it because they believe that a fundamental object must have no size, it must be a point. But that does not follow: lack of spatial extend and lack of substructure are independent concepts.

In fact, we can conceptually construct an infinitesimal object that is not elementary. Suppose we have dots moving on the surface of a sphere. Their motion can be expressed in terms of their polar and azimuthal angles. If we shrink the sphere, the dynamics and its description stays the same. We can collapse the whole sphere to a point: it is now a point-like object that is not elementary, as there is motion detectable along different angle of incidence.

Not only point-like objects don't have to be elementary, but it's unclear how can they be elementary. How can the whole universe be reduced to  elements that have no idea how many spatial dimensions are there? Mathematically a set of points by itself does not define the dimension of a space: $\mathbbm{R}^n$ is the same order of infinity for any $n$. One needs a topology to differentiate, which is an additional structure. It's the neighborhoods that define the dimensionality. If everything is just points, where does the topology come from?

If we abandon the idea of point-like objects of finite mass, we avoid these problems and get a picture more consistent across the different realms of physics. And we can also see that classical mechanics already has some features that are thought to be inherent to quantum mechanics.

@@

Suppose we have our classical electron described by a finite amount of classical material distributed over a single d.o.f. $\int \rho_\mathcal{c} dq \wedge dp = 1$ where $\rho_\mathcal{c}$ is the normalized distribution. In this context, we will call \emph{fragment} an infinitesimal part of our classical electron. The Shannon entropy associated with the distribution is finite as well. $I(\rho_\mathcal{c}) = - \int \rho_\mathcal{c} \log (\rho_\mathcal{c})  dq \wedge dp = I_0$. Intuitively, this value can be though as the amount of information required to identify a fragment within the distribution.\footnote{To be precise, Shannon's differential entropy is actually the number of bits required to identify an element of the given distribution relative to one of a uniform distribution of unitary range on all axis.}

As the state evolves into its final state, the amount of material remains the same: $\int \rho_{\hat{\mathcal{c}}} dq \wedge dp = \int \rho_{\mathcal{c}} dq \wedge dp$. The final entropy will also remain the same: $I(\rho_{\hat{\mathcal{c}}}) = I_0$. Mathematically because $-\rho_\mathcal{c} \log (\rho_\mathcal{c})$, like the distribution itself, is just another integrable function and therefore its integral is conserved by Hamiltonian evolution. Physically, because the amount of information required to identify a fragment given the distribution has to remain the same under deterministic and reversible evolution. Therefore, in the same way that each fragment is constrained to move through states at equal energy, the composite state is constrained to move through distributions at equal entropy.

We now ask the following question: once we fix the entropy, what is the relationship between the spread of the marginal distributions along $q$ and $p$? Using Lagrange multipliers, we fix the total amount of material to $1$ and the total entropy $I_0$ and find the distribution that minimizes the the product of the variance $\sigma_q^2 \sigma_p^2 = \int (q-\mu_q)^2 \rho \, dqdp \int (p-\mu_p)^2 \rho \, dqdp$.
\begin{align*}
L = &\int (q-\mu_q)^2 \rho \, dqdp \int (p-\mu_p)^2 \rho \, dqdp \\
&+ \lambda_1(\int \rho dqdp - 1)\\
&+ \lambda_2(- \int \rho \ln \rho \, dqdp - I_0)\\
\delta L = &\int \delta \rho [(q-\mu_q)^2 \sigma_p^2 + \sigma_q^2 (p-\mu_p)^2 \\ &+ \lambda_1 - \lambda_2 \ln \rho - \lambda_2 ] dqdp = 0 \\
\lambda_2 \ln \rho = &\lambda_1 - \lambda_2 + (q-\mu_q)^2 \sigma_p^2 + \sigma_q^2 (p-\mu_p)^2 \\
\rho = &e^{\frac{\lambda_1 - \lambda_2}{\lambda_2}}e^{\frac{(q-\mu_q)^2 \sigma_p^2}{\lambda_2}}e^{\frac{\sigma_q^2 (p-\mu_p)^2}{\lambda_2}}\\
\end{align*}
We solve the multipliers and have:
\begin{align*}
\rho = &\frac{1}{ 2 \pi \sigma_q \sigma_p} e^{-\frac{(q-\mu_q)^2}{2\sigma_q^2}} e^{-\frac{(p-\mu_p)^2}{2\sigma_p^2}} \\
I_0 = &\ln (2\sigma_q\sigma_p\pi) + 1
\end{align*}
We find that the distribution that minimizes the spread is the product of two independent Gaussians. Recall that in quantum mechanics, the Gaussian wave packet is the state that minimizes uncertainty: this is the classical analogue.

@@

Suppose, though, that while the evolution of the distribution of our electron electron as a whole is deterministc and reversible, each fragment is undergoing non deterministic motion. Suppose $\rho$ is a distribution at equilibrium of such motion. While each fragment moves randomly, the overall distribution $\rho$ stays the same: the non deterministic motion is effectively just reshuffling the fragments within the same distribution. The value of the entropy of the distribution, and therefore its spread, is set by the non deterministic motion of the particles: the stronger it is, the greater the entropy.

@@

If we assumed that all our classical electrons were affected by the same source of randomness, all fragments would exhibit a random walk of the same magnitude as defined by the information entropy. The spread of all distributions for the material would satisfy the same inequality. This gives us the classical uncertainty principle: $\sigma_q\sigma_p \geq \exp (I_\varnothing - 1) / 2 \pi $ where $I_\varnothing$ is the entropy associated with the source.

Such an assumption is not far fetched. The source of entropy could be identified in the environment (i.e. a minimum interaction that cannot be avoided as perfect isolation cannot be achieved) or the internal self interaction of the fragments (i.e. the evolution of the unstated part). Whatever it may be, if it characterized by an entropy comparable to the one of the distribution, that is we are trying to localize the electron better than the source can allow, the classical assumption breaks down as the random motion of each fragment is not negligible.

The generalization to multiple d.o.f. is outside of the scope of this work. Note, though, that this result is similar to the Gromov non-squeezing theorem and the symplectic camel, which are already generalized to multiple d.o.f.. It is the link with information theory that gives us further physical insight and is of particular interest given the strong connection between thermodynamics and Shannon's entropy.

@@

Consider the a planarian flatworm. If we cut it into multiple pieces, each part will grow into a new worm. Yet, the state of each worm is not the state of multiple worms put together. The planarian is divisible into multiple worms but it is not reducible to multiple worms. Consider a magnet. Its state is described, to a first approximation, by the position of the north and south poles. Yet, we cannot divide the magnet into a north and south pole. The magnet is reducible to the poles but is not divisible into the poles.

@@

Even more subtle is the difference between divisibility and decomposability, the idea that the object has parts. That is, a system is decomposable if there is a law $+ : \mathcal{S} \times \mathcal{S} \rightarrow \mathcal{S}$ to combine two states to get a third. While irreducibility says nothing on whether a quantum particle can be divided, it implicitly assumes that it can be decomposed. Irreducibility describes the relationship between the state of the whole and the state of the parts, therefore the existence of parts are implicit in the definition. But it also tells us that we are limited in what description we can give to these parts: we cannot isolate them as we want and act on them independently because if we were able to do that we could assign them an independent state.

Therefore, while we must be able to talk about parts of a quantum particles, not all questions will make sense. Can we ask what fraction of a quantum particle is in a particular region of space? Yes, since the particle can be spatially decomposed some amount of material may be inside the volume. Can we ask where that same fraction will be at a later time? No, since the particle is irreducible we cannot describe the evolution of a part.

Note that if the material is infinitesimally reducible, reducibility/decomposability/divisibility are all possible. This is why one may not dwell on their difference and use the word ``part" interchangeably in the three cases. Now that we assume irreducibility, it is better to use more precise language to avoid confusion. Therefore we will call \emph{fragments} the infinitesimal parts of the material that cannot be described independently. We will call \emph{components} the parts that the material can be decomposed into. A component will have a state associated with it, and therefore will always be a collection of fragments. As in the classical case, a component may be the difference between two states, therefore one should not imagine it always ``contained" in the composed system.

@@

The second feature is non locality. Suppose we have a quantum particle distributed in space. Suppose we exert a force only on a subregion of its distribution. As the particle is irreducible, we cannot interact with only part of it: we must also affect the part where no force is exerted. We cannot simply interact with the part at a specific region, as this would allow to assign a state to that part.

Wave particle duality is also accounted for. The quantum particle behaves like a wave in the sense that its parts form a distribution in space. Yet, it behaves as a single unit, as the state is irreducible and cannot be split.

The most striking feature, though, is that we can predict the functional form for particle decay over time. Suppose a quantum particle spontaneously decays into two or more particles. Such process cannot be deterministic as the number of d.o.f. increases. As the decay is spontaneous, it is not triggered by the environment (i.e. external state). It must be triggered by the internal dynamic of the fragments. As we assumed the internal dynamic is unknown, we cannot learn anything about it by looking at the product of the decay. In particular, all states have to have equal chance of decay or we could tell that the internal dynamic is more stable for some states. The probability of decay is constant for all states at all times. The probability of survival follows the exponential distribution.

@@

As for the classical material, the irreducible material is thought to be decomposable: each state can seen as the sum of different components (e.g. spatially distributed part) such that the evolution of the whole is the composition of their evolution. Therefore \ref{prop:abelian_group} applies and the state space $\Psi$ of an irreducible material is an abelian group.

Note, though, that the components are in general not independent. As the motion of each fragment is statistical in nature, it will combine as statistical variables do. Suppose $X$ and $Y$ are two random variables,
\begin{equation}\label{eqn:correlation}
\sigma^2_{X+Y} = \sigma^2_{X} + \sigma^2_{Y} + 2 \, \sigma_{X} \sigma_{Y} \rho_{X,Y}
\end{equation}
where $\sigma^2$ is the variance, $\sigma$ is the standard deviation, and $\rho_{X,Y}=COV(X,Y)/\sigma_{X} \sigma_{Y}$ is the Pearson correlation coefficient. As $-1\leq\rho_{X,Y}\leq+1$, the total variance can be greater or smaller then the sum of of the variance of the components. The strength of the chaotic motion of the fragments is something physically distinguishable as it contributes to the spread of the distribution. Therefore while the motion of each fragment is not itself physically distinguishable, it falls within the unstated part, the strength and correlation of the chaotic motion must be, they fall within the state.

How can we characterize the aspects of the unstated part that are captured by the state? We can look at the classical derivation for guidance. In that case, we looked at the set of transformation that increase/decrease the amount of material. We can think them as the transformations that change the unstated part of the system. As the classical material is homogeneous, permutations between classical particles in the same state $\mathcal{s}$ do not matter (i.e. the ordering is unstated). Increasing/decreasing the amount of material increases/decreases the number of permutations and therefore changes the unstated part of the system. For the irreducible material we follow the same strategy: we look for the transformations that affect the unstated part while keeping the overall state, the distribution, unchanged.

Suppose we have $\psi_1$ a state for an amount $a_m$ of irreducible material characterized by one d.o.f. The material will be spread over the configurations of the infinitesimal fragments. As the material is homogeneous, the material is spread evenly among such possible configurations. Counting fragment configurations is therefore equivalent to counting amount of material. Each fragment would be identified by a pair of independent random variables $(A, B)$, as a d.o.f. is two-dimensional. The shape of the distribution and the actual values are not physically distinguishable: these are just dummy variables we assign to count the fragments. We can choose A and B to be uniform distributions of interval $\sqrt{a_m}$, so that they identify a square of area $a_m$, the amount of material.

@@

The transformation $\tau(c)$ increases/decreases the area by the modulus $|c|^2=a^2+b^2$, which physically corresponds to the increase/decrease of amount of material. It also rotates the square by the phase $\theta_c \equiv arg(c) = arctan(a/b)$, which corresponds in a change in correlation: if $\theta_c = 0$, $C = |c| A$ are clearly correlated while if $\theta_c = \pi$, $C = - |c| A$ are clearly anti-correlated.

To better understand the link between phase and correlation, suppose $\psi$ is the state for an irreducible material. We can write $\tau(c) \psi = \tau(c_1 + c_2) \psi = \tau(c_1) \psi + \tau(c_2) \psi$ as the composition of two separate transformations. We have
\begin{equation}\label{eqn:correlation_state}
|c|^2=|c_1|^2 + |c_2|^2 + 2 |c_1||c_2|\cos(\theta_{c_2} - \theta_{c_1})
\end{equation}
Compare to \ref{eqn:correlation}. The material is spread evenly among fragment configurations. Also, since it is homogeneous, the ratio of fragment per configuration is the same. Therefore the amount of material is proportional to the strength of the chaotic motion of the fragment, which is measured by the variance. The increase/decrease in amount of material therefore combines like the variance of a statistical variable. The cosine of the phase difference takes the place of the Pearson correlation coefficient. A more meaningful name for the phase, then, would be \emph{correlation angle}. Also note how correlation for a single variable is a meaningless quantity: it is the correlation between two random variables that is significant. That is why in quantum mechanics phase differences have physical meaning, as it gives the correlation between two states, while absolute phases do not.

@@

Mathematically, the state space $\Psi$ for an irreducible material is a complex vector space. It is so because it is the state space for a decomposable material (i.e. an abelian group) where we can increase/decrease the amount and change the correlation of the internal stochastic motion (i.e. a set of transformation conveniently parametrized by a complex number).

>>>>

We will see that, as a first approximation, quantum particles are more suitably described by classical composite states (i.e. distributions in phase space) like any other object made of a finite amount of material. We'll also see that classical distributions already possess qualitative properties that are typically associated with quantum systems, including a classical analogue of the uncertainty principle.

@@

The first one is that quantum particles are not associated with a well defined value for position and momentum. If we misidentify the electron with a classical particle (i.e. the infinitesimal part of a classical material), this would indeed be worthy of note and puzzlement. But if the electron is a distribution, there is no issue as a phase space distribution does not have a single value for position or momentum. The earth is, so to speak, in both North America and Europe at the same time. In fact, based on what we saw before, any object of finite mass is a distribution over a region of phase space, and discontinuous distributions are ruled out by \ref{continuous_map}. An electron is an object of finite mass: why should it be any different? In fact, if one claimed an electron to be point-like, he would be claiming an electron is fundamentally different than all other physical objects of finite mass/energy (which are the only ones we experimentally work with). Shouldn't \emph{he} have the burden of proof?

@@

While this picture is quite reasonable, some people may reject it because they believe that a fundamental object must have no size, it must be a point. But that does not follow: lack of spatial extent and lack of substructure are independent concepts.

In fact, we can conceptually construct an infinitesimal object that is not elementary. Suppose we have dots moving on the surface of a sphere. Their motion can be expressed in terms of their polar and azimuthal angles. If we shrink the sphere, the dynamics and its description stays the same. We can collapse the whole sphere to a point: it is now a point-like object that is not elementary, as there is motion detectable along different angles of incidence.

Not only do point-like objects not have to be elementary, but it's unclear how they can be elementary. How can the whole universe be reduced to  elements that have no idea how many spatial dimensions are there? Mathematically a set of points by itself does not define the dimension of a space: $\mathbbm{R}^n$ is the same order of infinity for any $n$. One needs a topology to differentiate, which is an additional structure. It's the neighborhoods that define the dimensionality. If everything is just points, where does the topology come from?

If we abandon the idea of point-like objects of finite mass, we avoid these problems and get a picture more consistent across the different realms of physics. And we can also see that classical mechanics already has some features that are commonly ascribed to quantum mechanics.

@@

Suppose we have our classical electron described by a finite amount of classical material distributed over a single d.o.f. $\int \rho_\mathcal{c} dq \wedge dp = 1$ where $\rho_\mathcal{c}$ is the normalized distribution. In this context, we will call \emph{fragment} an infinitesimal part of our classical electron. The Shannon entropy associated with the distribution is finite as well. $I(\rho_\mathcal{c}) = - \int \rho_\mathcal{c} \log (\rho_\mathcal{c})  dq \wedge dp = I_0$. Intuitively, this value can be thought of as the amount of information required to identify a fragment within the distribution.\footnote{To be precise, Shannon's differential entropy is actually the number of bits required to identify an element of the given distribution relative to one of a uniform distribution of unitary range on all axes.}

As the state evolves into its final state, the amount of material remains the same: $\int \rho_{\hat{\mathcal{c}}} dq \wedge dp = \int \rho_{\mathcal{c}} dq \wedge dp$. The final entropy will also remain the same: $I(\rho_{\hat{\mathcal{c}}}) = I_0$. Mathematically, because $-\rho_\mathcal{c} \log (\rho_\mathcal{c})$, like the distribution itself, is just another integrable function and therefore its integral is conserved by Hamiltonian evolution. Physically, because the amount of information required to identify a fragment given the distribution has to remain the same under deterministic and reversible evolution. Therefore, in the same way that each fragment is constrained to move through states at equal energy, the composite state is constrained to move through distributions at equal entropy.

We now ask the following question: once we fix the entropy, what is the relationship between the spread of the marginal distributions along $q$ and $p$? Using Lagrange multipliers, we fix the total amount of material to $1$ and the total entropy $I_0$ and find the distribution that minimizes the product of the variance $\sigma_q^2 \sigma_p^2 = \int (q-\mu_q)^2 \rho_{\mathcal{c}} \, dqdp \int (p-\mu_p)^2 \rho_{\mathcal{c}} \, dqdp$.
\begin{align*}
L = &\int (q-\mu_q)^2 \rho_{\mathcal{c}} \, dqdp \int (p-\mu_p)^2 \rho_{\mathcal{c}} \, dqdp \\
&+ \lambda_1(\int \rho_{\mathcal{c}} dqdp - 1)\\
&+ \lambda_2(- \int \rho_{\mathcal{c}} \ln \rho_{\mathcal{c}} \, dqdp - I_0)\\
\delta L = &\int \delta \rho_{\mathcal{c}} [(q-\mu_q)^2 \sigma_p^2 + \sigma_q^2 (p-\mu_p)^2 \\ &+ \lambda_1 - \lambda_2 \ln \rho_{\mathcal{c}} - \lambda_2 ] dqdp = 0 \\
\lambda_2 \ln \rho_{\mathcal{c}} = &\lambda_1 - \lambda_2 + (q-\mu_q)^2 \sigma_p^2 + \sigma_q^2 (p-\mu_p)^2 \\
\rho_{\mathcal{c}} = &e^{\frac{\lambda_1 - \lambda_2}{\lambda_2}}e^{\frac{(q-\mu_q)^2 \sigma_p^2}{\lambda_2}}e^{\frac{\sigma_q^2 (p-\mu_p)^2}{\lambda_2}}\\
\end{align*}
We solve the multipliers and have:
\begin{align*}
\rho_{\mathcal{c}} = &\frac{1}{ 2 \pi \sigma_q \sigma_p} e^{-\frac{(q-\mu_q)^2}{2\sigma_q^2}} e^{-\frac{(p-\mu_p)^2}{2\sigma_p^2}} \\
I_0 = &\ln (2\pi\sigma_q\sigma_p) + 1
\end{align*}
We find that the distribution that minimizes the spread is the product of two independent Gaussians. Recall that in quantum mechanics, the Gaussian wave packet is the state that minimizes uncertainty: this is the classical analogue.

@@

Suppose, though, that while the evolution of the distribution of our electron as a whole is deterministc and reversible, each fragment is undergoing non deterministic motion. Suppose $\rho$ is a distribution at equilibrium of such motion. While each fragment moves randomly, the overall distribution $\rho$ stays the same: the non deterministic motion is effectively just reshuffling the fragments within the same distribution. The value of the entropy of the distribution, and therefore its spread, is set by the non deterministic motion of the particles: the stronger it is, the greater the entropy.

@@

If we assumed that any classical electron were affected by the same source of randomness, all fragments would exhibit a random walk of the same magnitude as defined by the information entropy. The spread of all distributions for the material would satisfy the same inequality. This gives us the classical uncertainty principle: $\sigma_q\sigma_p \geq \exp (I_\varnothing - 1) / 2 \pi $ where $I_\varnothing$ is the entropy associated with the source.

Such an assumption is not far fetched. The source of entropy could be identified in the environment (i.e. a minimum interaction that cannot be avoided as perfect isolation cannot be achieved) or the internal self interaction of the fragments (i.e. the evolution of the unstated part). Whatever it may be, if it is characterized by an entropy comparable to the one of the distribution, that is we are trying to localize the electron better than the source can allow, the classical assumption breaks down as the random motion of each fragment is not negligible.

The generalization to multiple d.o.f. is outside of the scope of this work. Note, though, that this result is related to the Gromov non-squeezing theorem and the symplectic camel, which are already generalized to multiple d.o.f.. It is the link with information theory that gives us further physical insight and is of particular interest given the strong connection between thermodynamics and Shannon's entropy.

@@

Consider a planarian flatworm. If we cut it into multiple pieces, each part will grow into a new worm. Yet, the state of each worm is not the state of multiple worms put together. The planarian is divisible into multiple worms but it is not reducible to multiple worms. Consider a magnet. Its state is described, to a first approximation, by the position of the north and south poles. Yet, we cannot divide the magnet into a north and south pole. The magnet is reducible to the poles but is not divisible into the poles.

@@

Even more subtle is the difference between divisibility and decomposability, the idea that the object has parts. That is, a system is decomposable if there is a law $+ : \mathcal{S} \times \mathcal{S} \rightarrow \mathcal{S}$ to combine two states to get a third. While irreducibility says nothing on whether a quantum particle can be divided, it implicitly assumes that it can be decomposed. Irreducibility describes the relationship between the state of the whole and the state of the parts, therefore the existence of parts is implicit in the definition. But it also tells us that we are limited in what description we can give to these parts: we cannot isolate them as we want and act on them independently because if we were able to do that we could assign them an independent state.

Therefore, while we must be able to talk about parts of a quantum particle, not all questions will make sense. Can we ask what fraction of a quantum particle is in a particular region of space? Yes, since the particle can be spatially decomposed some amount of material may be inside the volume. Can we ask where that same fraction will be at a later time? No, since the particle is irreducible we cannot describe the evolution of a part.

Note that if the material is infinitesimally reducible (i.e. the classical case), reducibility/decomposability/divisibility are all possible. This is why one may not dwell on their difference and use the word ``part" interchangeably in the three cases. Now that we assume irreducibility, it is better to use more precise language to avoid confusion. Therefore we will call \emph{fragments} the infinitesimal parts of the material that cannot be described independently. We will call \emph{components} the parts that the material can be decomposed into. A component will have a state associated with it, and therefore will always be a collection of fragments. As in the classical case, a component may be the difference between two states, therefore one should not imagine it always ``contained" in the composed system.

@@

The second feature is non locality. Suppose we have a quantum particle distributed in space. Suppose we exert a force only on a subregion of its distribution. As the particle is irreducible, we cannot interact with only part of it: we must also affect the part where no force is exerted. We cannot simply interact with the part at a specific region, as this would allow us to assign a state to that part.

Wave-particle duality is also accounted for. The quantum particle behaves like a wave in the sense that its parts form a distribution in space. Yet, it behaves as a single unit, as the state is irreducible and cannot be split.

The most striking feature, though, is that we can predict the functional form for particle decay over time. Suppose a quantum particle spontaneously decays into two or more particles. Such process cannot be deterministic as the number of d.o.f. increases. As the decay is spontaneous, it is not triggered by the environment (i.e. external state). It must be triggered by the internal dynamics of the fragments. As we assumed the internal dynamics is unknown, we cannot learn anything about it by looking at the product of the decay. In particular, all states have to have equal chance of decay or we could tell that the internal dynamics is more stable for some states. The probability of decay is constant for all states at all times. The probability of survival follows the exponential distribution.

@@

As for the classical material, the irreducible material is thought to be decomposable: each state can be seen as the sum of different components (e.g. spatially distributed parts) such that the evolution of the whole is the composition of their evolution. Therefore \ref{prop:abelian_group} applies and the state space $\Psi$ of an irreducible material is an abelian group.

Note, though, that the components are in general not independent. As the motion of each fragment is statistical in nature, it will combine as statistical variables do. Suppose $X$ and $Y$ are two random variables,
\begin{equation}\label{eqn:correlation}
\sigma^2_{X+Y} = \sigma^2_{X} + \sigma^2_{Y} + 2 \, \sigma_{X} \sigma_{Y} \rho_{X,Y}
\end{equation}
where $\sigma^2$ is the variance, $\sigma$ is the standard deviation, and $\rho_{X,Y}=COV(X,Y)/\sigma_{X} \sigma_{Y}$ is the Pearson correlation coefficient. As $-1\leq\rho_{X,Y}\leq+1$, the total variance can be greater or smaller than the sum of the variance of the components. The strength of the chaotic motion of the fragments is something physically distinguishable as it contributes to the spread of the distribution. Therefore while the motion of each fragment is not itself physically distinguishable, it falls within the unstated part, the strength and correlation of the chaotic motion must be, they fall within the state.

How can we characterize the aspects of the unstated part that are captured by the state? We can look at the classical derivation for guidance. In that case, we looked at the set of transformations that increase/decrease the amount of material. We can think of them as the transformations that change the unstated part of the system. As the classical material is homogeneous, permutations between classical particles in the same state $\mathcal{s}$ do not matter (i.e. the ordering is unstated). Increasing/decreasing the amount of material increases/decreases the number of permutations and therefore changes the unstated part of the system. For the irreducible material we follow the same strategy: we look for the transformations that affect the unstated part while keeping the overall state, the distribution, unchanged.

Suppose we have $\psi_1$ a state for an amount $a_m$ of irreducible material characterized by one d.o.f. The material will be spread over the configurations of the infinitesimal fragments. As the material is homogeneous, it is spread evenly among such possible configurations. Counting fragment configurations is therefore equivalent to counting the amount of material. Each fragment would be identified by a pair of independent random variables $(A, B)$, as a d.o.f. is two-dimensional. The shape of the distribution and the actual values are not physically distinguishable: these are just dummy variables we assign to count the fragments. We can choose A and B to be uniform distributions of interval $\sqrt{a_m}$, so that they identify a square of area $a_m$, the amount of material.

@@

The transformation $\tau(c)$ increases/decreases the area by the modulus $|c|^2=a^2+b^2$, which physically corresponds to the increase/decrease of amount of material. It also rotates the square by the phase $\theta_c \equiv arg(c) = arctan(a/b)$, which corresponds to a change in correlation: if $\theta_c = 0$, $C = |c| A$ are clearly correlated while if $\theta_c = \pi$, $C = - |c| A$ are clearly anti-correlated.

To better understand the link between phase and correlation, suppose $\psi$ is the state for an irreducible material. We can write $\tau(c) \psi = \tau(c_1 + c_2) \psi = \tau(c_1) \psi + \tau(c_2) \psi$ as the composition of two separate transformations. We have
\begin{equation}\label{eqn:correlation_state}
|c|^2=|c_1|^2 + |c_2|^2 + 2 |c_1||c_2|\cos(\theta_{c_2} - \theta_{c_1})
\end{equation}
Compare to \ref{eqn:correlation}. The material is spread evenly among fragment configurations. Also, since it is homogeneous, the ratio of fragments per configuration is the same. Therefore the amount of material is proportional to the strength of the chaotic motion of the fragment, which is measured by the variance. The increase/decrease in amount of material therefore combines like the variance of a statistical variable. The cosine of the phase difference takes the place of the Pearson correlation coefficient. A more meaningful name for the phase, then, would be \emph{correlation angle}. Also note how correlation for a single variable is a meaningless quantity: it is the correlation between two random variables that is significant. That is why in quantum mechanics phase differences have physical meaning, as it gives the correlation between two states, while absolute phases do not.

@@

Mathematically, the state space $\Psi$ for an irreducible material is a complex vector space. It is so because it is the state space for a decomposable material (i.e. an abelian group) where we can increase/decrease the amount and change the correlation of the internal stochastic motion (i.e. a set of transformations conveniently parametrized by a complex number).
